---
title: "STAT-731 Recitation 2"
author: "K. Tyler Wilcox"
date: "September 2, 2016"
output:
  beamer_presentation:
    highlight: zenburn
    incremental: true
    theme: "metropolis"
---

Preliminaries
========================================================

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

- Install `R` Packages
    - `distr`
- Why Use Computer Simulation?

Probability Mass Function
========================================================

- $X$ is a discrete random variable
- $f(x)$ is a function
- $f(x) = Pr[X = x]$ for $x \in X$ is the *probability distribution* or *probability mass function* (pmf) of $X$
- $f(x)$ must satisfy:
    - $f(x) = Pr[X = x] \ge 0, \forall x \in X$
    - $\sum_{x \in X} f(x) = \sum_{x \in X} Pr[X = x] = 1$

Discrete Distributions: More Coins
========================================================

- Let's revisit our favorite pastime: coin flips
- $\mathcal{X} = \{ H, T \}$
- $Pr[X = H] = 0.7$
- $Pr[X = T] = 1 - Pr[X = H] = 0.3$
- $f(x) = {n \choose x} \theta ^ {x} (1 - \theta) ^ {n - x}, x \ge 0, \theta \ge 0$
- Is this a valid pmf?
- What distribution is this?

Visualizing the Binomial Distribution
========================================================

- Let's sample from $\mathcal{X}$ ten times

```{r, message = FALSE, fig.height = 4.5}
library(distr)
outcome = distr::Binom(size = 10, prob = 0.7)
plot(outcome)
```

***

- What if $Pr[X = H] = 0.3$?

```{r, fig.height = 4.5}
outcome = distr::Binom(size = 10, prob = 0.3)
plot(outcome)
```

***

- What if $Pr[X = H] = 0.5$?

```{r, fig.height = 4.5}
outcome = distr::Binom(size = 10, prob = 0.5)
plot(outcome)
```

Tossing Coins
========================================================

- First, let's write a simple `R` function that flips a coin
- We will set the number of coin flips with the `nflips` argument
- We can even make the coin fair or unfair with the `weights` argument


```{r}
flip_coin = function(nflips = 1,
                     weights = c(0.5, 0.5)) {

  sequence = sample(x       = c("H", "T"),
                    size    = nflips,
                    replace = TRUE,
                    prob    = weights)
  return(sequence)
}
```

Tossing Coins: What Does a Fair Coin Look Like?
========================================================

- Let's flip a fair two-sided coin (Heads or Tails) once
- We will use our `flip_coin` function
- `set.seed()` ensures that we can rerun the same code and get the same result
even though we are using a pseudo-random number generator

```{r}
set.seed(333)
sequence = flip_coin(nflips = 1)
print(sequence)
```

Tossing Coins: Two Tosses
========================================================

- Do we have a fair coin based on what we saw?
- Let's flip a fair two-sided coin (Heads or Tails) twice

```{r}
sequence = flip_coin(nflips = 2)
print(sequence)
```

- Is this a fair sequence?

Tossing Coins: Three Tosses
========================================================

- What about 3 tosses?

```{r}
sequence = flip_coin(nflips = 3)
print(sequence)
```

Tossing Coins: 10,000 Tosses
========================================================
- What does 3 tosses look like?

```{r, echo = FALSE}
library(pacman)
p_load(ggplot2, scales, dplyr)

ggsequence = sequence %>%
  data.frame() %>%
  tbl_df()
names(ggsequence) = "Outcome"

sequence2 = flip_coin(nflips = 10000) %>%
  data.frame() %>%
  tbl_df()
names(sequence2) = "Outcome"

ggplot(ggsequence) +
  geom_bar(aes(x = Outcome, y = ..count.. / sum(..count..))) +
  scale_y_continuous(name = "Relative Frequency", labels = percent,
                     limits = c(0.0, 1.0)) +
  ggtitle("Three Tosses") +
  theme_minimal()
```

***

- What about 10,000 tosses?

```{r, echo = FALSE}
ggplot(sequence2) +
  geom_bar(aes(x = Outcome, y = ..count.. / sum(..count..))) +
  scale_y_continuous(name = "Relative Frequency", labels = percent,
                     limits = c(0.0, 1.0)) +
  ggtitle("10,000 Tosses") +
  theme_minimal()
```

Tossing Coins: Probability as Relative Frequency
========================================================

- Effectively, the probability of an event is defined by the relative frequency
of the event as the number of independent trials approaches infinity

- For our fair coin toss, the probability of "Heads" is:
$$Pr\left[ H \right] = \lim_{n \to \infty} \frac{1}{n}\sum_{i = 1} ^ n I(H)$$

- Let's use the function `sample()` to compute the probability of seeing "Heads"

`sample()`
========================================================

- `sample()` is a function that draws a random sample from a sample space
    - `x`: a vector containing the elements of the sample space
        - `sample(x = c("H", "T"))`
    - `size`: a non-negative integer specifying the sample size to draw
        - `sample(x = c("H", "T"), size = 10)`
    - `replace`: `TRUE` or `FALSE`; sample with or without replacement
        - default is `replace = FALSE`
        - `sample(x = c("H", "T"), size = 10, replace = TRUE)`
    - `prob`: a vector of probabilities for each element of the sample space `x`
        - default is `1 / length(x)`
        - `sample(x = c("H", "T"), size = 10, replace = TRUE, prob = c(0.75, 0.25))`

Tossing Coins: Probability of "Heads" by Simulation
========================================================

- Let $n$ be very large (approximately infinite)
- `nheads` will count the number of times a "Heads" is obtained

```{r}
n = 100000
nheads = 0
for (i in seq(1:n)) {
  nheads = nheads + ifelse(
    test = sample(x = c("H", "T"))[1] == "H",
    yes  = 1,
    no   = 0)
}
```

***

- We obtained $`r sprintf("%1.f", nheads)`$ "Heads" out of $`r sprintf("%1.f", n)`$
tosses of a fair coin
- Therefore, $Pr\left[H\right] = `r sprintf("%.4f", nheads / n)`$
- How close are we to the theoretical probability?

Tossing Coins: A Rigged Coin
========================================================

- $Pr[H] = 0.75$
- $Pr[T] = 1 - Pr[H] = 0.25$

```{r}
scams = flip_coin(nflips = 40,
                  weights = c(0.75, 0.25))
table(scams)
```

- We probably don't have a fair coin when
`r sprintf("%.2f", (table(scams)[1] / length(scams) * 100))`% of the flips
landed on "Heads"

Tossing Coins: Probability of Rigged "Heads" by Simulation
========================================================

- Let $n$ be very large (approximately infinite)
- `nheads` will count the number of times a "Heads" is obtained

```{r}
n = 100000
nheads = 0
for (i in seq(1:n)) {
  nheads = nheads + ifelse(
    test = sample(x = c("H", "T"),
                  prob = c(0.75, 0.25))[1] == "H",
    yes  = 1,
    no   = 0)
}
```

***

- We obtained $`r sprintf("%1.f", nheads)`$ "Heads" out of $`r sprintf("%1.f", n)`$ tosses of a fair coin
- Therefore, $Pr\left[H\right] = `r sprintf("%.4f", nheads / n)`$
- How close are we to the theoretical probability?

Rolling Dice
========================================================

- We spent a good amount of time simulating binomial events
- Can we simulate probabilities for larger spaces?

***

- Let's roll two 6-sided fair dice:
$$\mathcal{S} = \{1, 2, 3, 4, 5, 6\} \times \{1, 2, 3, 4, 5, 6\}$$
- What is the probability of rolling the same number on both dice?

```{r}
n = 10000
outcome = 0
for (i in seq(1:n)) {
  outcome = outcome + ifelse(
    test = sample(x = 1:6)[1] == sample(x = 1:6)[1],
    yes = 1,
    no = 0)
}
```

***

- We empirically obtain $`r sprintf("%1.f", outcome)`$ out of
$`r sprintf("%1.f", n)`$ rolls
- Our empirical estimate of $Pr[f_1 = f_2] = `r sprintf("%.4f", outcome / n)`$
- How does this compare to our theoretical expectation?

Empirical Exploration of an Arbitrary Sample Space
========================================================

- Let $\mathcal{S} = \{-2, 0, 1, 3\}$
- $Pr[X = -2] = Pr[X = 0] = Pr[X = 1] = Pr[X = 3] = 0.25$
- What is the size (cardinality) of $\mathcal{S}$?
- Let's compute $Pr[X = -2]$ empirically using relative frequencies

```{r}
n = 10000
outcome = 0
for (i in seq(1:n)) {
  outcome = outcome + ifelse(
    test = sample(x = c(-2, 0, 1, 3),
                  size = 1, replace = TRUE) == -2,
    yes = 1,
    no = 0)
}
```

***

- We observed $`r sprintf("%1.f", outcome)`$ events where $X = -2$ out of
$`r sprintf("%1.f", n)`$ samples
- Our empirical estimate of $Pr[X = -2] = `r sprintf("%.4f", outcome / n)`$
- How does this compare to our theoretical expectation?

Empirical Exploration of an Arbitrary Sample Space
========================================================

- Let $\mathcal{S} = \{-2, 0, 1, 3\}$
- $Pr[X = -2] = Pr[X = 0] = Pr[X = 1] = Pr[X = 3] = 0.25$
- What is the probability that X is non-negative?

```{r}
set.seed(333)
n = 10000
outcome = 0
for (i in seq(1:n)) {
  outcome = outcome + ifelse(
    test = sample(x = c(-2, 0, 1, 3),
                  size = 1, replace = TRUE) >= 0,
    yes = 1,
    no = 0)
}
```

***

- We observed $`r sprintf("%1.f", outcome)`$ events where $X \ge 0$ out of
$`r sprintf("%1.f", n)`$ samples
- Our empirical estimate of $Pr[X \ge 0] = `r sprintf("%.4f", outcome / n)`$
- How does this compare to our theoretical expectation?

A Programming Optimization Aside
========================================================

- It is worth mentioning that `for()` loops are highly inefficient in `R`
- Let's redo the last example and replace the `for()` loop with one vectorized
line of code

```{r}
n = 10000
set.seed(333)
outcome2 = length(which(sample(x = c(-2, 0 , 1, 3),
                  size = n, replace = TRUE) >= 0))
all.equal(outcome, outcome2)
```

- In general, vectorization will be preferable when possible in `R`

Combinatorics in `R`
========================================================

- Consider an alphabet:
$$A = \{ a, b, c, d, e \}$$
- Suppose that for this alphabet, only words of length 2, 3, 4, and 5 are valid
- Further, suppose that letters can appear more than once per word
- Finally, suppose that the order of letters is irrelevant (i.e., "ab" = "ba")

Alphabet A: How Many Two Letter Words Are There?
========================================================

```{r, message = FALSE}
library(prob)

A = c('a', 'b', 'c', 'd', 'e')
n2words = prob::nsamp(n = length(A), k = 2,
                      replace = TRUE, ordered = FALSE)
```

- We have `r sprintf("%1.f", n2words)` two-letter words

Alphabet A: How Many Three Letter Words Are There?
========================================================

```{r}
A = c('a', 'b', 'c', 'd', 'e')
n3words = prob::nsamp(n = length(A), k = 3,
                      replace = TRUE, ordered = FALSE)
```

- We have `r sprintf("%1.f", n3words)` possible three-letter words

Alphabet A: How Many Four Letter Words Are There?
========================================================

```{r}
A = c('a', 'b', 'c', 'd', 'e')
n4words = prob::nsamp(n = length(A), k = 4,
                      replace = TRUE, ordered = FALSE)
```

- We have `r sprintf("%1.f", n4words)` four-letter words

Alphabet A: Two-Letter Words in the Language?
========================================================

```{r}
dictionary = NULL
w2 = urnsamples(x = A, size = 2,
                replace = TRUE, ordered = FALSE)
for (i in seq(1:nrow(w2))) {
  newword = ''
  for (j in seq(1:2)) {
    newword = paste0(newword, as.character(w2[i, j]))
  }
  dictionary = c(dictionary, newword)
}
```

```{r, echo = FALSE}
cat(dictionary)
```

Alphabet A: Three-Letter Words in the Language?
========================================================

```{r}
dictionary = NULL
w3 = urnsamples(x = A, size = 3,
                replace = TRUE, ordered = FALSE)
for (i in seq(1:nrow(w2))) {
  newword = ''
  for (j in seq(1:3)) {
    newword = paste0(newword, as.character(w2[i, j]))
  }
  dictionary = c(dictionary, newword)
}
```

***

```{r, echo = FALSE}
cat(dictionary[1:10], "\n")
cat(dictionary[11:20], "\n")
cat(dictionary[21:30], "\n")
cat(dictionary[31:35], "\n")
```

Alphabet A: Four-Letter Words in the Language?
========================================================

```{r}
dictionary = NULL
w4 = urnsamples(x = A, size = 4,
                replace = TRUE, ordered = FALSE)
for (i in seq(1:nrow(w2))) {
  newword = ''
  for (j in seq(1:4)) {
    newword = paste0(newword, as.character(w2[i, j]))
  }
  dictionary = c(dictionary, newword)
}
```

***

```{r, echo = FALSE}
cat(dictionary[1:10], "\n")
cat(dictionary[11:20], "\n")
cat(dictionary[21:30], "\n")
cat(dictionary[31:40], "\n")
cat(dictionary[41:50], "\n")
cat(dictionary[51:60], "\n")
cat(dictionary[61:70], "\n")
```

Wrapping Up
========================================================

- `R` and RStudio
- Theoretical and empirical (simulation-based) probabilities
- When to use a theoretical vs. empirical approach
- Questions?
